Classification and Regression Trees and Beyond

### 2020-06-30 pkg/partykit/inst/ULGcourse-2020/slides_tree.Rnw
streamline R code
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3020 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 80c1c0b24be6d1e76e628c9e970c540b56484e29

<<Titanic-data>>=
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
@

@@

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
ct <- ctree(Survived ~ Gender + Age + Class, data = ttnc,
            control = ctree_control(alpha = 0.01))
plot(ct)
@
\end{center}
\end{frame}

@@

\end{frame}
\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
data("CPS1985", package = "AER")
plot(ctree(log(wage)~education+experience+age+ethnicity+gender+union,data=CPS1985,
           control = ctree_control(alpha = 0.01)))
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education \ | \ experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
data("CPS1985", package = "AER")
plot(lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985))
@
\end{center}
\end{frame}

@@

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Gender)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Gender, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Gender, data = ttnc)
@
\end{minipage}

@@

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Gender)
@
\end{frame}

@@

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Age)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Age, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Age, data = ttnc)
@
\end{minipage}

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Age)
@
\end{frame}

@@

\vspace{0.5cm}

\begin{minipage}{0.5\linewidth}
\vspace{-2.2cm}
<<echo=TRUE, results=verbatim>>=
table(ttnc$Survived, ttnc$Class)
@
\end{minipage}
\begin{minipage}{0.45\linewidth}
<<echo=TRUE, eval=FALSE>>=
plot(Survived~Class, data = ttnc)
@
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(Survived~Class, data = ttnc)
@
\end{minipage}

<<echo=TRUE, results=verbatim>>=
chisq.test(ttnc$Survived, ttnc$Class)
@
\end{frame}

% 
% \begin{frame}[fragile]
% \frametitle{Split variable selection} 
% {\bf Compare pairwise associations} (Titanic): 
% Based on independence tests, e.g., $\chi^2$-test for categorical variables.
% \vspace{0.4cm}
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived~Gender, data = ttnc)
% @
% 
% \vspace{0.2cm}
% 
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived~Age, data = ttnc)
% @
% 
% \vspace{0.7cm}
% 
% $\Rightarrow$ Select the split variable showing the lowest $p$-value.
% 
% \end{frame}

@@

\begin{minipage}{0.4\linewidth}
<<echo=TRUE, results=verbatim>>=
cor(log(CPS1985$wage), 
    CPS1985$education)
@
<<echo=TRUE, eval=FALSE>>=
plot(log(wage)~education, 
     data = CPS1985)
@
\end{minipage}
\hfill
\begin{minipage}{0.5\linewidth}
%\vspace{-0.2cm}
%\setkeys{Gin}{width=0.5\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
plot(log(wage)~education, data = CPS1985)
@
\end{minipage}

<<echo=TRUE, results=verbatim>>=
cor.test(log(CPS1985$wage), CPS1985$education)
@

@@

\vspace{0.2cm}

<<echo=TRUE, results=verbatim>>=
tapply(log(CPS1985$wage), CPS1985$gender, summary)
@
\begin{minipage}{0.37\linewidth}
\vspace{-2cm}
<<echo=TRUE, eval=FALSE>>=
boxplot(log(wage)~gender, 
        data = CPS1985)
@
\end{minipage}
\hfill
\begin{minipage}{0.55\linewidth}
\vspace{-0.7cm}
%\setkeys{Gin}{width=0.4\linewidth}
<<echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
boxplot(log(wage)~gender, data = CPS1985)
@
\end{minipage}
\end{frame}

\vspace{0.2cm}

<<echo=TRUE, results=verbatim>>=
oneway.test(log(wage)~gender, data = CPS1985, var.equal = TRUE)
@
\end{frame}

@@

<<split_point_search>>=

evalsplit <- function(sp, type = c("rmse","teststatistic")){

  if(type == "rmse"){
    d1 <- CPS1985[CPS1985$education <= sp,]
    d2 <- CPS1985[CPS1985$education > sp,]

    y1 <- mean(log(d1$wage))
    y2 <- mean(log(d2$wage))

    rmse1 <- sum(sqrt((log(d1$wage) - y1)^2))
    rmse2 <- sum(sqrt((log(d2$wage) - y2)^2))

    rmse <- rmse1 + rmse2

    return(rmse)
  }

  if(type == "teststatistic"){

    itest <- independence_test(log(wage) ~ factor(education <= sp), data = CPS1985) 
    pval <- as.numeric(pvalue(itest))
    teststatistic <- statistic(itest)
    return(list(pval = pval,
                teststatistic = teststatistic))
  }
}

evalpoints <- sort(unique(CPS1985$education))
evalpoints <- evalpoints[-length(evalpoints)]
plotpoints_rmse <- numeric(length = length(evalpoints))
for(j in 1: length(plotpoints_rmse))
  plotpoints_rmse[j] <- evalsplit(evalpoints[j], type = "rmse")

plotpoints_statistic <- plotpoints_pval <- numeric(length = length(evalpoints))
for(k in 1: length(plotpoints_statistic)){
  test <- evalsplit(evalpoints[k], type = "teststatistic")
  plotpoints_statistic[k] <- test$teststatistic
  plotpoints_pval[k] <- test$pval
}

@

@@

%\setkeys{Gin}{width=0.8\textwidth}
\begin{minipage}{0.49\linewidth}
<<plot_split_point_search_rmse, fig=TRUE, width = 4.5, height = 4>>=
plot(x = evalpoints, y = plotpoints_rmse, type = 'o',
     xlab = "education", ylab = "RMSE")
@
\end{minipage}
\hfill
\begin{minipage}{0.49\linewidth}
<<plot_split_point_search_statistic, fig=TRUE, width = 4.5, height = 4>>=
plot(x = evalpoints, y = plotpoints_statistic, type = 'o',
     xlab = "education", ylab = "Two-sample test statistic")

# plot(x = evalpoints, y = plotpoints_pval, type = 'o',
#      xlab = "education", ylab = "twosample-teststatistic",
#      ylim = c(0,0.1))

@
\end{minipage}
\end{frame}

>>>>

<<data>>=
data("CPS1985", package = "AER")
data("Titanic", package = "datasets")
ttnc <- as.data.frame(Titanic)
ttnc <- ttnc[rep(1:nrow(ttnc), ttnc$Freq), 1:4]
names(ttnc)[2] <- "Gender"
@

@@

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Titanic:  \quad $Survived \; \sim  \; Gender + Age + Class$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<ctree_Titanic, echo=FALSE, fig=TRUE, width=11>>=
ct_ttnc <- ctree(Survived ~ Gender + Age + Class, data = ttnc, alpha = 0.01)
plot(ct_ttnc)
@
\end{center}
\end{frame}

@@

\end{frame}
\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education + experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.99\textwidth}
<<ctree_Wages1985, echo=FALSE, fig=TRUE, width=13>>=
ct_cps <- ctree(log(wage) ~ education + experience + age + ethnicity + gender + union, data = CPS1985, alpha = 0.01)
plot(ct_cps)
@
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Motivation}
{\small Wages:  \quad $log(wage) \; \sim  \; education \ | \ experience + age + ethnicity + gender + union$}
\begin{center}
\setkeys{Gin}{width=0.9\textwidth}
<<lmtree_Wages1985, echo=FALSE, fig=TRUE, width=11>>=
mob_cps <- lmtree(log(wage) ~ education | experience + age + ethnicity + gender + union, data = CPS1985)
plot(mob_cps)
@
\end{center}
\end{frame}

@@

\begin{minipage}[t]{0.43\linewidth}
<<survival-gender, echo=TRUE, eval=FALSE>>=
plot(Survived ~ Gender, data = ttnc)
@
\vspace{-0.7cm}
<<survival-gender-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
<<survival-gender>>
@
\end{minipage}
\hfill
\pause
\begin{minipage}[t]{0.5\linewidth}
<<echo=TRUE, results=verbatim>>=
xtabs(~ Survived + Gender, data = ttnc)
@
\end{minipage}

@@

<<echo=TRUE, results=verbatim>>=
chisq.test(xtabs(~ Survived + Gender, data = ttnc))
@

@@

\begin{minipage}[t]{0.43\linewidth}
<<survival-age, echo=TRUE, eval=FALSE>>=
plot(Survived ~ Age, data = ttnc)
@
\vspace{-0.7cm}
<<survival-age-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
<<survival-age>>
@
\end{minipage}
\hfill
\pause
\begin{minipage}[t]{0.5\linewidth}
<<echo=TRUE, results=verbatim>>=
xtabs(~ Survived + Age, data = ttnc)
@
\end{minipage}

<<echo=TRUE, results=verbatim>>=
chisq.test(xtabs(~ Survived + Age, data = ttnc))
@

@@

\begin{minipage}[t]{0.43\linewidth}
<<survival-class, echo=TRUE, eval=FALSE>>=
plot(Survived ~ Class, data = ttnc)
@
\vspace{-0.7cm}
<<survival-class-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
<<survival-class>>
@
\end{minipage}
\hfill
\pause
\begin{minipage}[t]{0.5\linewidth}
<<echo=TRUE, results=verbatim>>=
xtabs(~ Survived + Class, data = ttnc)
@
\end{minipage}

<<echo=TRUE, results=verbatim>>=
chisq.test(xtabs(~ Survived + Class, data = ttnc))
@

% 
% \begin{frame}[fragile]
% \frametitle{Split variable selection} 
% {\bf Compare pairwise associations} (Titanic): 
% Based on independence tests, e.g., $\chi^2$-test for categorical variables.
% \vspace{0.4cm}
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Gender, data = ttnc)
% @
% 
% \vspace{0.2cm}
% 
% <<echo=TRUE, results=verbatim>>=
% chisq_test(Survived ~ Age, data = ttnc)
% @
% 
% \vspace{0.7cm}
% 
% $\Rightarrow$ Select the split variable showing the lowest $p$-value.
% 
% \end{frame}

@@

\begin{minipage}[t]{0.39\linewidth}
<<logwage-education, echo=TRUE, eval=FALSE>>=
plot(log(wage) ~ education,
  data = CPS1985)
@
\vspace{-0.7cm}
\setkeys{Gin}{width=\textwidth}
<<logwage-education-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
<<logwage-education>>
@
\end{minipage}
\hfill
\pause
\begin{minipage}[t]{0.59\linewidth}
<<echo=TRUE, eval=FALSE>>=
cor.test(~ log(wage) + education, data = CPS1985)
@
<<echo=FALSE, results=verbatim>>=
out <- capture.output(cor.test(~ log(wage) + education, data = CPS1985))
out <- gsub("alternative hypothesis: ", "alternative hypothesis:\n", out, fixed = TRUE)
writeLines(out)
@
\end{minipage}

@@

\begin{minipage}[t]{0.39\linewidth}
<<logwage-gender, echo=TRUE, eval=FALSE>>=
plot(log(wage) ~ gender,
  data = CPS1985)
@
\vspace{-0.7cm}
\setkeys{Gin}{width=\textwidth}
<<logwage-gender-plot, echo=FALSE, fig=TRUE, width=4.5, height=3.4>>=
<<logwage-gender>>
@
\end{minipage}
\hfill
\pause
\begin{minipage}[t]{0.59\linewidth}
<<echo=TRUE, eval=FALSE>>=
t.test(log(wage) ~ gender, data = CPS1985)
@
<<echo=FALSE, results=verbatim>>=
out <- capture.output(t.test(log(wage) ~ gender, data = CPS1985))
out <- gsub("alternative hypothesis: ", "alternative hypothesis:\n", out, fixed = TRUE)
writeLines(out)
@
\end{minipage}

@@

<<split_point_search>>=
evalsplit <- function(sp) {
  m <- lm(log(wage) ~ factor(education <= sp), data = CPS1985)
  s <- independence_test(log(wage) ~ factor(education <= sp), data = CPS1985, teststat = "quadratic") 
  c(
    rss = deviance(m),
    loglik = logLik(m),
    logpval = pchisq(statistic(s), df = 1, log = TRUE, lower.tail = FALSE),
    teststatistic = statistic(s)
  )
}  
sp <- head(sort(unique(CPS1985$education)), -1)
eval <- sapply(sp, evalsplit)
@

@@

<<plot_split_point_search1, fig=TRUE, width = 7.5, height = 5>>=
par(mar = c(5, 3, 1, 3))
plot(sp, eval["loglik",], type = "n", xlab = "education", ylab = "", axes = FALSE)
abline(v = 13, lty = 2, lwd = 1.5)
lines(sp, eval["loglik", ], type = "o", col = 2, lwd = 1.5)
axis(2, col.ticks = 2, col.axis = 2)
par(new = TRUE)
plot(sp, eval["teststatistic",], type = "o",
  col = 4, lwd = 1.5,
  xlab = "", ylab = "", axes = FALSE)
axis(4, col.ticks = 4, col.axis = 4)
axis(1)
axis(1, at = 13)
box()
legend("topleft", c("Log-likelihood", "Test statistic"), col = c(2, 4), lwd = 1.5, pch = 1, bty = "n")
@

<<plot_split_point_search2, fig=TRUE, width = 7.5, height = 5>>=
par(mar = c(5, 3, 1, 3))
plot(sp, eval["rss",], type = "n", xlab = "education", ylab = "", axes = FALSE)
abline(v = 13, lty = 2, lwd = 1.5)
lines(sp, eval["rss", ], type = "o", col = 2, lwd = 1.5)
axis(2, col.ticks = 2, col.axis = 2)
par(new = TRUE)
plot(sp, eval["logpval",], type = "o",
  col = 4, lwd = 1.5,
  xlab = "", ylab = "", axes = FALSE)
axis(4, col.ticks = 4, col.axis = 4)
axis(1)
axis(1, at = 13)
box()
legend("bottomleft", c("Residual sum of squares", "Log-p-value"), col = c(2, 4), lwd = 1.5, pch = 1, bty = "n")
@
